# An Introduction to NFL Analytics and the R Programming Language

```{r loading-library, include = FALSE}

library(tidyverse)
```

## Introduction

It probably seems odd to jump right into coding and visualization in Chapter 1, while placing information about learning the basics of the `tidyverse` in Chapter 2. But there is good reason I adopted this pedagogical approach in this book. As explained by Hadley Wickham and Garrett Grolemund in their outstanding book [*R for Data Science*](https://r4ds.had.co.nz/index.html)*,* the process of reading in and then cleaning data is not exactly the most exciting part of doing analytics. As evidence suggest, early excitement and integration into a topic increases the likelihood of following up and learning the boring material.

Because of this, I follow the approach of Wickham and Grolemund and provide you data that is already imported and, for the most part, tidied and ready to go.

::: callout-important
I am assuming, while you may not have a full grasp of the `tidyverse` yet, **that you do currently have base R and RStudio installed**. If you do not, more detailed instructions are provided in Chapter 2. If you would rather jump right into this material, you can download base R and RStudio at the following links. Once both are installed, you can return to this point in the chapter to follow along.

To download/install base R: [cran.rstudio.com](https://cran.rstudio.com/)\
To download/install RStudio: [RStudio Desktop](https://posit.co/download/rstudio-desktop/) (scroll to bottom of page for Mac options)
:::

In this chapter, we will use data I prepared from the 2021 NFL season. **add more about what the chapter does here**.

## 2021 NFL Completion Leaders

At the conclusion of the 2021 NFL season, Tom Brady - of the Tampa Bay Buccaneers - led the league in completed pass attempts with 485, well above Justin Herbert's tally of 443 completions for second place. However, as is the case with many NFL statistics, there is much context that can be added to even an otherwise "simple statistic" such as pass completions.

For example, one may be interested in seeing the progression of completion throughout the season to witness how the "race to the top" changed from week to week. And, of course, all completions are not created equal. Was Tom Brady doing a "nickel and dime" routine down the field, thus picking up easier completions than, for example, a QB slinging it 15+ yards at a higher rate than Brady? Or what about a quarterback's time to throw? It seems logical to conclude that an under performing offensive line, providing limited time for a QB to make reads and get rid of the ball, would impact multiple parts of a QB's passing game.

To start exploring some of these questions and assumptions, lets pull the prepared data from this book's GitHub using the following code:

```{r reading-example-data, include = TRUE, output = FALSE}

pbp <- readr::read_csv("https://raw.githubusercontent.com/bcongelio/nfl-analytics-with-r-book/origin/example_data/complete_passes.csv")

```

::: callout-note
The `readr` package is part of the `tidyverse`. Please make sure to run `library(tidyverse` to load both.
:::
